Analysis-ready VCF at Biobank scale using Zarr

### 2024-11-21 paper.tex

% Middle
\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
\author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284

@@

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[7]{Wellcome Sanger Institute}
\affil[8]{Genomics England}
\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
  \midrule
  /call\_PL & int16 & 6.12 TiB & 12.0 & 92.75\% \\
  /call\_genotype & int8 & 282.45 GiB & 26.0 & 4.18\% \\
  /call\_genotype\_mask & bool & 42.15 GiB & 180.0 & 0.62\% \\
  /variant\_DP4 & int32 & 26.24 GiB & 2.1 & 0.39\% \\
  /variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.19\% \\
  /variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.19\% \\
  /variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.19\% \\
  /variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.19\% \\
  /variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.19\% \\
  /variant\_quality & float32 & 12.84 GiB & 1.1 & 0.19\% \\
  /variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.19\% \\
  /variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.18\% \\
  /variant\_position & int32 & 9.65 GiB & 1.4 & 0.14\% \\
  /variant\_AC & int16 & 7.42 GiB & 2.8 & 0.11\% \\
  /variant\_DP & int32 & 6.71 GiB & 2.1 & 0.10\% \\
  /variant\_allele & object & 5.37 GiB & 21.0 & 0.08\% \\
  /variant\_AN & int16 & 3.05 GiB & 2.3 & 0.05\% \\
  /call\_genotype\_phased & bool & 2.03 GiB & 1800.0 & 0.03\% \\
  /variant\_filter & bool & 1.46 GiB & 2.4 & 0.02\% \\
  /variant\_MQ & int8 & 489.83 MiB & 7.3 & 0.01\% \\
  \bottomrule
\end{tabular}
\end{table}

@@

\subsection{Funding}
JK acknowledges the Robertson Foundation and 
NIH (research grants HG011395 and HG012473). 
JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
TM acknowledges funding from The New Zealand Institute for Plant \& Food
Research Ltd Kiwifruit Royalty Investment Programme.

% \subsection{Author's Contributions}

@@

>>>>

% Middle
\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
\author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
\author[10]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315

@@

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[7]{Wellcome Sanger Institute}
\affil[8]{Genomics England}
\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}
\affil[10]{Department of Cell and Molecular Biology, National
  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
  Uppsala University, Uppsala, Sweden}

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
  \midrule
  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
  \bottomrule
\end{tabular}
\end{table}

@@

\subsection{Funding}
JK acknowledges the Robertson Foundation and 
NIH (research grants HG011395 and HG012473). 
JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
TM acknowledges funding from The New Zealand Institute for Plant \& Food
Research Ltd Kiwifruit Royalty Investment Programme.
PU was supported by the SciLifeLab \& Wallenberg Data Driven Life
Science Program, Knut and Alice Wallenberg Foundation (grants: KAW
2020.0239 and KAW 2017.0003), and by the National Bioinformatics
Infrastructure Sweden (NBIS) at SciLifeLab
% \subsection{Author's Contributions}

@@

Computation for the Spruce case study were enabled by resources
provided by the National Academic Infrastructure for Supercomputing in
Sweden (NAISS), partially funded by the Swedish Research Council
through grant agreement no. 2022-06725.
